---
title: Language learnability variability
author: "Molly Lewis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
---
  
***
***

```{r setup, echo = F, warning = F, message = F}
rm(list = ls())

library(knitr)
library(ggplot2)
library(langcog)
library(plyr)
library(dplyr)
library(tidyr)
library(broom)
library(lme4)
library(rworldmap)
library(factoextra)
library(grid)
library(gridExtra)
library(ggrepel)

opts_chunk$set(cache = F, warning = F, message = F)

mapTheme = theme(plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.line = element_blank(), 
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank())

removeOutlier <- function(dataCol, num_deviations) {
  boolean.outlier = abs(dataCol) > mean(dataCol, na.rm = T) +
                    (num_deviations*sd(dataCol, na.rm = T))
  df = bind_cols(as.data.frame(boolean.outlier),
                 as.data.frame(dataCol))
  cleaned.dataCol = ifelse(df$boolean.outlier, NA, df$dataCol)
  cleaned.dataCol
}

most.frequent.level = function (x){
    mf = names(which.max(table(x)))
    return(mf)
}

N_EXCLUDE_SDS = 10
CORR_THRESHOLD = .8
```

## Variable descriptions
 * Lupyan & Dale (2010): Demographic variables and syntactic complexity
    + log population size (pop_log)
    + area (area)
    + number of bordering countries (n.neigbors)
    + mean temperature (mean.temp)
    + variability in temperature (sd.temp)
    + total precipitation (sum.precip)
    + variability in precipition (sd.precip)
    + growing season (growing.season)
    + latitude (lat)
    + longitude (lon)
    + morpholigical complexity (morphological.complexity)
 * Bentz, et al. (2015): L2 learners and lexical diversity
    + number of first language speakers (n.L1.speakers)
    + number of second language speakers (n.L2.pseakers)
    + ratio of L2 to L1 (ratio.L2.L1)
    + parameters of Zipf-Mandelbrot’s law for lexical diversity (scaled.LDT.ZM)
    + Shannon entropy of lexical diversity (scaled.LDT.H)
    + Type-token ratio as measure of lexical diversity (scaled.LDT.TTR)
 * Atkinson (2011): Phonemic diversity and best-fitting distance from origin
    + vowel diversity (normalized.vowel.diversity)
    + consonant diversity (normalized.consonant.diversity)
    + tone diversity (normalized.tone.diversity)
    + phoneme diversity (normalized.phoneme.diversity)
    + estimated distance for language origin (distance.from.origin)
 * Lewis & Frank (under review): Complexity bias
    + complexity bias, correlation between length and complexity (complexity.bias)
    + bias partially out spoken frequency (p.complexity.bias)
    + bias for only monosyllabic words (mono.complexity.bias)
    + bias for only open class words (open.complexity.bias)
 * Futrell, Mahowald, & Gibson (2015): Dependency length
    + mean observed dependency length in characters (mean.dependency.length)
 * Pellegrino, Coupe, & Marsico (2015): Information density
     + information.density
     + syllable.rate
     + information.rate
 * Moran, McCloy, & Wright (2012): Phonemic inventory
    + number of phonemes (n.phonemes)
    + number of cononsants (n.consonants)
    + number of vowels (n.vowels)
    + number of sonorants (n.sonorants)
    + number of obsruents (n.obsruents)
    + number of monophthongs  (n.monophthongs)
    + number of quality-only monophthong constrasts (n.qual.monophthongs)
    + number of tones (n.tones)
 * Wichmann, et al. (2013): Mean word length
    + mean length in number of characters (mean.length)
 * Luniewska, et al. (2015): Age of acquisition (aoa)
    + mean adult-judged age of acquistion (mean.aoa)

## Read in datasets
Language codes (from WALS) -- WALS, ISO, ascii-name. These are used to merge across datasets. 
```{r read in datasets}
codes = read.csv("../data/language_codes.csv") %>%
  select(language, WALS, ISO) %>%
  mutate(ISO = unlist(lapply(strsplit(as.character(ISO),
                                      ","),function(x) x[1])))
# in cases where there are two ISO codes, takes first
```

Lupyan & Dale (2010): Demographic variables and syntactic compelxity
```{r}
wichmann_ISOS = read.csv("../data/ISO_mappings_from_wichmann.csv", 
                         stringsAsFactors=F)[-1] %>%
  unique() 

ld = read.table("../data/lupyan_2010.txt", fill = T, 
                header = T, sep = "\t", na.strings = "*") %>%
  left_join(codes, c("walsCode" = "WALS")) %>%
  mutate(lang = tolower(lang)) %>%
  left_join(wichmann_ISOS, by = "lang") %>% # use wichmann ISOS for some missing
  mutate(ISO = as.factor(ifelse(is.na(ISO.x), ISO.y, ISO.x))) %>%
  select(-ISO.y, -ISO.x) %>%
  filter(!is.na(ISO)) %>%
  filter(ISO != "")

# rename columns
ld = rename(ld, pop_log = logpop2, 
         mean.temp = aveTemp,
         n.neighbors = numNeighbors,
         sum.precip = sumPrecip,
         sd.temp = sdTemp,
         sd.precip = sdPrecip,
         lang.family = langFamily,
         lang.genus = langGenus,
         native.country = nativeCountry,
         native.country.area = nativeCountryArea,
         growing.season = growingSeason) 

# get means across language
ld.demo = ld %>%
  group_by(ISO) %>%
  summarise_each(funs(mean(., na.rm = TRUE)), 
                 c(8:9, 16:121))  %>%
  select(1:3, 93:95, 100:101, 103:105, 108)

# add in data family
fams = ld %>%
  group_by(ISO) %>%
  filter(row_number() == 1) %>%
  select(ISO, lang.family, lang.genus, native.country,
         native.country.area, language)  

ld.demo = left_join(fams, ld.demo, by = "ISO") %>% ungroup()
```

WALS features of complexity
```{r}
# get variables to include 
qual = ld %>%
  select(18:103, 124) %>%
  mutate_each(funs(as.factor)) %>%
  group_by(ISO) %>%
  summarise_each(funs(most.frequent.level)) %>%
  mutate_each(funs(as.factor)) 

ld_feature_names = read.csv("../data/lupyan_2010_all_feature_mappings.csv") 
qualVarNames = intersect(names(qual), ld_feature_names$WALS.feature.name)
qual = select(qual,which(is.element(names(qual), c("ISO", qualVarNames))))

# remap factor levels to complexity values (0,1)
# note two variables that are reported in the paper (NICPOC[59] and CYSIND[39] are missing in the data)

for (i in 1:length(qualVarNames)){
  thisVarLabs = ld_feature_names[ld_feature_names$WALS.feature.name == qualVarNames[i],]
  old = thisVarLabs$ld.level.label
  if (is.na(old)) {print("NA!!!")}
  new = thisVarLabs$ld.complexity
  col_i = grep(qualVarNames[i], colnames(qual))
  qual[,col_i] = mapvalues(as.matrix(qual[,col_i]), 
                           from = as.character(old),
                           to = as.character(new), warn_missing = TRUE)
}

ld.complexity = qual %>%
  mutate_each(funs(as.numeric), -ISO) %>%
  gather(variable, complexity.level, 2:28) %>%
  group_by(ISO) %>%
  summarise(morphological.complexity = sum(complexity.level))

ld.demo.qual = left_join(ld.demo, ld.complexity)
```

Bentz, et al. (2015): L2 learners and lexical diversity.Note here that we are only looking at languages from the UDHR corpus. The other corpora have smaller languages only.
```{r, eval = F}
bentz = read.csv("../data/bentz_2015.csv") %>%
  gather(temp, LDT, starts_with("LDT")) %>% 
  unite(temp1, measure, temp, sep = ".") %>% 
  spread(temp1, LDT) %>%
  filter(text == "UDHR") %>%
  select(iso_639_3, RatioL2, a.LDTscaled, 
         H.LDTscaled, TTR.LDTscaled, Stock,
         Region, L1_speakers, L2_speakers, TTR.LDT) %>%
  rename(ISO = iso_639_3,
         n.L1.speakers = L1_speakers,
         n.L2.speakers = L2_speakers,
         ratio.L2.L1 = RatioL2,
         stock = Stock,
         region = Region,
         scaled.LDT.TTR = TTR.LDTscaled,
         scaled.LDT.ZM = a.LDTscaled,
         scaled.LDT.H = H.LDTscaled) 

# n_speakers taken directly from ethnologue website by ML for missing languages that were high-frequency across the datasets
ethno_sup = read.csv("../data/supplementary_enthnologue.csv")
bentz = full_join(bentz, ethno_sup) %>%
        mutate(ISO = as.factor(ISO))
```

Atkinson (2011)
```{r, eval = F}
atkinson = read.csv("../data/atkinson_2011.csv") %>%
  select(normalized.vowel.diversity,
         normalized.consonant.diversity,	
         normalized.tone.diversity,
         normalized.phoneme.diversity, ISO, 
         distance.from.origin) %>%
  mutate(ISO = as.factor(unlist(lapply(strsplit(as.character(ISO),
                                      " "),function(x) x[1])))) %>%
  distinct(ISO) 
```

Lewis & Frank (under review): Complexity Bias
```{r, eval = F}
# complexity bias
cb = read.csv("../data/lewis_2015.csv") %>%
  rename(complexity.bias = corr,
         p.complexity.bias = p.corr,
         mono.complexity.bias = mono.cor,
         open.complexity.bias = open.cor) %>%
  left_join(codes, by="language") %>%
  select(-X.1, -X, -lower.ci, -upper.ci, -checked,
         -mean.length) %>%
  distinct(ISO) %>%
  filter(language != "english") %>% # english is an outlier in this dataset because norms were colelcted in english
  select(-language, -WALS)
```

Futrell, Mahowald, & Gibson (2015): Dependency length
```{r, eval = F}
dl = read.csv("../data/futrell_2015.csv") %>%
  left_join(codes, by="language") %>%
  select(-language, -WALS, -fixed.random.baseline.slope, 
         -observed.slope, -m_rand) %>%
  rename(mean.dependency.length = m_obs) %>%
  filter(!is.na(ISO))
```

Pellegrino, Coupe, & Marsico (2015): Information density
```{r, eval = F}
uid = read.csv("../data/pellegrino_2015.csv")  %>%
  left_join(codes, by="language") %>%
  select(-language, -WALS)
```

Moran, McCloy, & Wright (2012): Phonemic inventory
```{r, eval = F}
phoneme = read.csv("../data/moran_2012.csv") %>%
  select(ISO, pho, con, vow, son, 
         obs, mon, qua, ton) %>%
  rename(n.phonemes = pho,
         n.consonants = con,
         n.vowels = vow,
         n.sonorants = son,
         n.obstruents = obs,
         n.monophthongs = mon,
         n.qual.monophthongs = qua,
         n.tones = ton)
```

Wichmann, et al. (2013): Mean word length
```{r, eval = F}
# note is this IPA? - fix this!
# [308 low frequency language have no ISO, leaving 4421 (TOTAL:4729)]
ml.raw = read.csv("../data/wichmann_2013.csv") %>% 
  select(1,1:109) %>%
  gather(word,translation,I:name) %>%
  mutate(nchar = unlist(
    lapply(
      lapply(
        strsplit(
          gsub("[[:space:]]", "", translation) ,
          ","), 
        nchar), mean))) %>%
  filter(translation != "")

# subset to only those words in the swadesh list (n = 40)
swadesh.words  = ml[ml$ISO == "gwj", "word"] 
ml.raw = ml.raw %>%
  filter(is.element(word, swadesh.words))

ml = ml.raw %>%
  group_by(ISO) %>%
  summarize(mean.length = mean(nchar, na.rm = T)) %>%
  filter(ISO != "")

#more_ISOs = ml.raw %>%
#  select(ISO, names) %>%
#  mutate(names = tolower(names))
```

Luniewska, et al. (2015): Age of acquistion (Aoa)
```{r, eval = T}
aoa.raw = read.csv("../data/luniewska_2015.csv", 
                   header = T, fileEncoding = "latin1")

aoa = aoa.raw %>%
  gather(language_aoa, aoa, grep("aoa", 
                                 names(aoa.raw))) %>%
  mutate(language_aoa = tolower(unlist(lapply(strsplit(
    as.character(language_aoa),"_"), function(x) x[1])))) %>%
  select(-3:-27) %>%
  rename(language = language_aoa) %>%
  left_join(codes, by="language") %>%
  mutate(language = as.factor(language)) %>%
  group_by(ISO) %>%
  summarize(mean.aoa = mean(aoa, na.rm = T)) %>%
  filter(!is.na(ISO))
```

Merge data frames together by ISO
```{r merge data, eval = F}
d = full_join(ld.demo.qual, cb, by = "ISO") %>%
  full_join(bentz, by = "ISO") %>%
  full_join(phoneme, by = "ISO") %>%
  full_join(ml, by = "ISO") %>%
  full_join(dl, by = "ISO") %>%
  full_join(uid, by = "ISO") %>%
  full_join(aoa, by = "ISO") %>%
  full_join(atkinson, by = "ISO") %>%
  mutate(ISO = as.factor(ISO)) %>%
  filter(!is.na(native.country)) # this is important to have for the mixed effects models

```

load data
```{r}
# write.csv(d, "../data/langLearnVar_data.csv")
d = read.csv("../data/langLearnVar_data.csv")[,-1]
```

## Clean data
Remove outliers and check for normality
```{r clean data, fig.width = 15, fig.height = 10}
d.clean = d %>%
  mutate_each(funs(removeOutlier(.,N_EXCLUDE_SDS)), c(-ISO, -lang.family, 
              -native.country, -native.country.area, 
         -lang.genus, -language, -stock, -region))

d.clean %>%
  select(-ISO, -lang.family, -native.country, -native.country.area, 
         -lang.genus, -language, -stock, -region) %>%
  gather(var, value) %>%
  ggplot(aes(sample = value)) + 
  stat_qq(na.rm = T, size = .2) + 
  facet_wrap( ~ var, scales = "free") +
  theme_bw()
```

The following variables look right-skewed. 
```{r}
NN_vars = c("area", "perimeter", "n.neighbors", "sd.precip",
           "ratio.L2.L1", "n.L1.speakers", "n.L2.speakers",
           "n.phonemes", "n.consonants","n.vowels", "n.sonorants",
           "n.obstruents", "n.monophthongs", "n.qual.monophthongs",
           "n.tones")
```

Take the log of these variables, and check for normality again.
```{r, fig.width = 15, fig.height = 10}
d.clean = d.clean %>%
  mutate_each(funs(log = log(.), dummy =  sum(.)),
              one_of(NN_vars)) %>%
  select(-contains("dummy")) %>%
  select(-one_of(NN_vars))

d.clean %>%
  select(-ISO, -lang.family, -native.country, -native.country.area, 
         -lang.genus, -language, -stock, -region) %>%
  gather(var, value) %>%
  ggplot(aes(sample = value)) + 
  stat_qq(na.rm = T, size = .2) + 
  facet_wrap( ~ var, scales = "free") +
  theme_bw()
```

Check for normality of variables using Shapiro-Wilk Normality Test
```{r}
is.na(d.clean) <- sapply(d.clean, is.infinite)

normality.test = d.clean %>%
  summarise_each(funs(unlist(tidy(shapiro.test(.))[2])),
                 c(-ISO, -lang.family, -native.country,
                   -native.country.area,
                     -lang.genus, -language, -stock, -region)) %>%
  gather(variable, shapiro.p.value) %>%
  mutate(normal = ifelse(shapiro.p.value > .05, TRUE, FALSE)) %>%
  summary()
```
By this test, only 11 variables normal. But, I think this is the best we can do (?). 

Look at histograms of all variables. 
```{r histograms, fig.height = 9, fig.width = 14}
demo_vars = c("lat", "lon", "perimeter_log", "n.neighbors_log",
              "mean.temp", "sum.precip",
              "sd.temp", "growing.season", "pop_log", "area_log",
              "n.neighbors_log", "sd.precip_log", "ratio.L2.L1_log",
              "n.L1.speakers_log",
              "n.L2.speakers_log", "distance.from.origin")

lang_vars = c(setdiff(names(d.clean), demo_vars))[-c(1:6, 22,23)]

d.clean %>%
  gather(variable, num, lat:length(d.clean))  %>%
  filter(variable != "stock", variable != "region") %>%
  mutate(var_type = ifelse(is.element(variable, demo_vars),
                           "demo", "lang"),
         num = as.numeric(as.character(num))) %>%
  arrange(var_type) %>% 
  mutate(variable = factor(variable, variable)) %>%
  ggplot(aes(x = num, fill = var_type)) + 
    geom_histogram(position = "identity") +
    facet_wrap( ~ variable, scales = "free") + 
    theme_bw()
```

Number of languages we have data, for each variable
```{r, fig.height = 6}
d.clean %>%
  gather(variable, num, lat:length(d.clean))  %>%
  filter(variable != "stock", variable != "region") %>%
  group_by(variable) %>%
  summarize(counts = length(which(is.na(num) == F)))  %>%
  mutate(var_type = ifelse(is.element(variable, demo_vars),
                           "demo", "lang")) %>%
  mutate(counts_trunc = ifelse(counts > 150, 150, counts))  %>%
  arrange(var_type) %>% 
  mutate(variable = factor(variable, variable)) %>%
  ggplot(aes(y=counts_trunc, x = variable, fill = var_type)) + 
    geom_bar(stat = "identity") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.position = "none") +
    ylab("Number of languages (truncated at 150)") +
    ggtitle("Number languages per measure")
```

Look at number of languages intersecting each variable 
```{r n_languages, fig.height = 10, fig.width = 10}
arealControl_vars = c("stock", "region", "lang.family", "lang.genus",
                      "native.country", "native.country.area")

d.clean %>%
  gather(lang_measure, lang_value, 
         which(is.element(names(.), c(lang_vars, arealControl_vars)))) %>%
  group_by(lang_measure) %>%
  gather(pop_measure, pop_value,
         which(is.element(names(.), demo_vars))) %>%
  group_by(lang_measure, pop_measure) %>%
  mutate(both_not_na = !is.na(lang_value) & !is.na(pop_value)) %>%
  summarise(n_languages = length(which(both_not_na == TRUE))) %>%
  ggplot(aes(pop_measure, lang_measure, group = lang_measure)) +
      geom_tile(aes(fill = n_languages)) + 
      geom_text(aes(fill = n_languages, label = n_languages)) +
      scale_fill_gradient(low = "white", high = "red")  +
      xlab("population variables") + 
      ylab("language variables") + 
      ggtitle("Number languages between variables") +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
     
## Relationship between language and demographic variables
### Fits controling for family in a mixed effect model
```{r model fits with all variables (lmer), fig.width=10, fig.height = 12}
# write.csv(d.clean,"../data/langLearnVar_data_clean.csv")
# d.clean = read.csv("../data/langLearnVar_data_clean.csv")[-1]

# these are the variables we care about for the paper
demo_vars_crit = c("n.neighbors_log",
              "mean.temp", "sum.precip", "sd.temp",
              "pop_log", "area_log", "n.neighbors_log",
              "sd.precip_log", "ratio.L2.L1_log")

lang_vars_crit = c("p.complexity.bias","TTR.LDT",
                   "mean.dependency.length", "mean.length",
                   "n.consonants_log", "n.vowels_log",
                   "morphological.complexity")

d.scatter.fam = d.clean %>% 
  select(pop_log, ratio.L2.L1_log,
         n.neighbors_log, area_log, mean.temp, sd.temp, sum.precip,
         sd.precip_log, morphological.complexity,TTR.LDT, p.complexity.bias,
         mean.length, n.consonants_log, n.vowels_log,
         lang.family, native.country) %>%
  gather(lang_measure, lang_value,
         which(is.element(names(.), lang_vars_crit))) %>%
  group_by(lang_measure) %>%
  gather(pop_measure, pop_value,
         which(is.element(names(.), demo_vars_crit))) 


is.na(d.scatter.fam) <- sapply(d.scatter.fam, is.infinite) 
d.scatter.fam <- filter(d.scatter.fam, !is.na(pop_value),
                        !is.na(lang_value))

# reorder factor levels for facet plots below (affects order in other dfs)
d.scatter.fam$lang_measure <- factor(d.scatter.fam$lang_measure,
                                    levels = c("morphological.complexity",
                                             "p.complexity.bias", "TTR.LDT",
                                             "mean.length", "n.consonants_log",
                                             "n.vowels_log"))

d.scatter.fam$pop_measure <- factor(d.scatter.fam$pop_measure,
                                    levels = c('pop_log', 'ratio.L2.L1_log',
                                               'n.neighbors_log', 'area_log',
                                               'mean.temp', 'sd.temp',
                                               'sum.precip', 'sd.precip_log' ))

# get model fits
d.model.fits = d.scatter.fam %>%
  group_by(lang_measure, pop_measure) %>%
  do(tidy(lmer(lang_value ~ pop_value + (pop_value|lang.family) +
                 (1|native.country), data=.))) 

line.fits = d.model.fits %>%
        filter(term == "pop_value" | term == "(Intercept)") %>%
        select(-std.error, -statistic, -group) %>%
        spread(term, estimate) %>%
        rename(s = pop_value, 
               i = `(Intercept)`)


sigs = d.model.fits %>%
  filter(term == "pop_value") %>%
  mutate(sig.col = ifelse(statistic > 1.96, "pos",
                          ifelse(statistic < -1.96, "neg",
                                 "none"))) %>%
  mutate(pop_value = .1, lang_value = .1) %>% # this is a hack
  ungroup
```

Paper plot #1
```{r}
facet_pop_names <- c(
  pop_log="Population \n(log)",
  ratio.L2.L1_log="L2:L1 \n(log)",
  n.neighbors_log="N. Ling. \nNeighbors (log)",
  area_log="Area \n(log)",
  mean.temp="Temperature \n Mean",
  sd.temp="Temperature \nSD",
  sum.precip="Precipitation \nSum",
  sd.precip_log="Precipitation \nSD (log)"
)

facet_lang_names <- c(
  morphological.complexity="Morphosyntactic \nComplexity",
  TTR.LDT="Lexical \nDiversity",
  p.complexity.bias="Complexity \nBias",
  mean.length="Word \nLength",
  n.consonants_log="N. Consonants \n(log)",
  n.vowels_log="N. Vowels \n(log)"
)
```

```{r, eval = F}
#pdf("../papers/cogsci2016/figs/plot1.pdf", width = 18, height = 14)
ggplot(d.scatter.fam, aes(x = pop_value, y = lang_value)) +
  facet_grid(lang_measure ~ pop_measure, 
             scales = "free", 
             labeller = labeller(pop_measure = facet_pop_names,
                                lang_measure = facet_lang_names)) +
  geom_rect(data = sigs, aes(fill = sig.col), 
          xmin = -Inf, xmax = Inf,
          ymin = -Inf, ymax = Inf, alpha = 0.2) +
  geom_point(size = 1.1, color = "grey48") +
  geom_abline(line.fits, mapping = aes(slope = s, intercept = i), 
              size = 2, color = "black") +
  scale_fill_manual(values = c( "mediumblue", "grey99", "red1")) +
  xlab("\nEnvironmental variables") +
  ylab("Language variables\n") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.title = element_text(size = 21), 
        strip.text= element_text(size = 17))
#dev.off()
```

### Replicate Lupyan and Dale (2010) complexity score plot (Fig. 3)
```{r Lupyan and Dale complexity plot}
ld.demo.qual %>%
  mutate(morphological.complexity.factor =
           as.factor(morphological.complexity)) %>%
  group_by(morphological.complexity.factor) %>%
  multi_boot(column="pop_log",
              summary_groups = c("morphological.complexity.factor"),
              statistics_functions = c("mean", "ci_lower","ci_upper")) %>%
  mutate(morphological.complexity =  
           as.numeric(as.character(morphological.complexity.factor))) %>%
  ggplot() +
  geom_pointrange(aes(x= morphological.complexity.factor,
                      y = mean, ymin = ci_lower, ymax = ci_upper)) + 
  geom_smooth(method = "lm",aes(morphological.complexity-5, mean)) +
  # theres a weird offset because complexity is a factor?
  ylab("Log population") + 
  xlab("Morphological complexity") +
  ggtitle("Lupyan and Dale (2010) Fig. 3 reproduction") +
  theme_bw() 
```

## Principle Component Analysis
###  PCA with demographic variables
prep by checking for correlations
```{r}
# get demographic variables only (excluding L2 because intersections contains only 40 languages)
d.demo =  d.clean %>%
  select(lat, mean.temp, sd.temp, sum.precip, sd.precip_log, 
         area_log, pop_log, n.neighbors_log) %>% 
  mutate(lat = abs(lat))

is.na(d.demo) <- sapply(d.demo, is.infinite) 

demo.corr = cor(d.demo, use = "pairwise.complete")
abs(demo.corr) > CORR_THRESHOLD

# exclude lat (correlated with mean.temp, sd.precip_log) [threshold: > .8]
```

PCA
```{r PCA with demographic variables}
d.demo =  select(d.demo, -lat)

# do pca
pca.demo = prcomp(na.omit(d.demo), scale = TRUE)
# look at variance explained
fviz_eig(pca.demo, addlabels = TRUE, 
               linecolor ="red", geom = "line") +
  theme_bw()

barplot(pca.demo$sdev/pca.demo$sdev[1])
pca.demo = prcomp(na.omit(d.demo), scale = TRUE, tol = .6)

# plot loadings
pcs.demo <- cbind(names(d.demo),
                    gather(as.data.frame(pca.demo$rotation), 
                           pc, value))
names(pcs.demo)[1] = "variable"
 
pcs.demo$variable = factor(pcs.demo$variable, levels = pcs.demo$variable[1:8]) 

# order variables
pcs.demo$magnitude = ifelse(abs(pcs.demo$value)>.2, TRUE, FALSE)
ggplot(pcs.demo) +
  geom_bar(aes(x = variable, y = value,
               fill = magnitude), stat="identity") +
  facet_wrap(~pc) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  + 
       scale_fill_manual(name = "Magnitude", 
                       values = c("grey","red1")) +
  ggtitle("PCA loadings")

# merge PCAs with full dataset for model fits
d.demo[!is.na(d.demo)] = 0
d.demo[is.na(d.demo)] = 1
good_is = which(rowSums(d.demo) == 0)
d.pca.demo = d.clean[good_is,] %>% cbind(pca.demo$x)

# rename PCs
names(d.pca.demo)[names(d.pca.demo) == "PC1"] = "hot"
names(d.pca.demo)[names(d.pca.demo) == "PC2"] = "small"
```

Maps of principle components for demographic variable by language
```{r principle components by language (maps), fig.height = 4, fig.width = 8}
ggplot(d.pca.demo) +   
  borders("world", colour="gray50", fill="gray50") +
  geom_point(aes(x = lon, y=lat, 
               color = hot), size=3) +
  scale_colour_gradient(high="green", low = "white") +
  ggtitle("Hot (PC1)") + 
  mapTheme

ggplot(d.pca.demo) +   
  borders("world", colour="gray50", fill="gray50") +
  geom_point(aes(x = lon, y=lat, 
               color = small), size=3) +
  scale_colour_gradient(high="blue", low = "white") +
  ggtitle("Small (PC2)") + 
  mapTheme
```

Model fits using demographic PCA with all language variables
```{r model fits with PCA for demo, fig.width=8, fig.height = 10}
demo_vars_crit2 = c("hot", "small")

lang_vars_crit2 = c("morphological.complexity",
                    "p.complexity.bias","TTR.LDT",
                    "mean.length", "mean.aoa",
                    "n.consonants_log", "n.vowels_log")

d.scatter.fam = d.pca.demo %>% 
  select(n.consonants_log, n.vowels_log, 
         mean.length, p.complexity.bias, TTR.LDT,
         morphological.complexity,
         lang.family, native.country, small, hot) %>%
  gather(lang_measure, lang_value, 
         which(is.element(names(.), lang_vars_crit2))) %>%
  group_by(lang_measure) %>%
  gather(pop_measure, pop_value,
         which(is.element(names(.), demo_vars_crit2)))

is.na(d.scatter.fam) <- sapply(d.scatter.fam, is.infinite) 
d.scatter.fam <- filter(d.scatter.fam, !is.na(pop_value),
                        !is.na(lang_value))

# get model fits
d.model.fits = d.scatter.fam %>%
  group_by(lang_measure, pop_measure) %>%
  do(tidy(lmer(lang_value ~ pop_value + 
                 (pop_value|lang.family) + 
                 (1|native.country),
               data=.)))

line.fits = d.model.fits %>%
          filter(term == "pop_value" | term == "(Intercept)") %>%
  select(-std.error, -statistic, -group) %>%
  spread(term, estimate) %>%
  rename(s = pop_value, 
        i = `(Intercept)`)

sigs = d.model.fits %>%
  filter(term == "pop_value") %>%
  mutate(sig.col = ifelse(statistic > 1.96, "pos",
                          ifelse(statistic < -1.96, "neg",
                                 "none"))) %>%
  mutate(pop_value = .1, lang_value = .1) %>% # this is a hack
  ungroup

ggplot(d.scatter.fam, aes(x = pop_value, y = lang_value)) +
  geom_rect(data = sigs, aes(fill = sig.col), 
        xmin = -Inf, xmax = Inf,
        ymin = -Inf, ymax = Inf, alpha = 0.2) +
  geom_point(size = .3) +
  geom_abline(line.fits, mapping = aes(slope = s, intercept = i), 
              size = 1, color = "green") +
  facet_grid(lang_measure~pop_measure, scales = "free") +
  scale_fill_manual(values = c( "mediumblue", "grey99","red1")) +
  theme_bw() +
  xlab("Demographic variables") +
  ylab("Language variables") +
  theme(legend.position = "none")
```

### PCA with language variables
prep by looking at correlations
```{r}
d.lang =  d.clean %>%
  select(morphological.complexity, scaled.LDT.H, 
         mean.length, n.consonants_log, n.vowels_log) 
is.na(d.lang) <- sapply(d.lang, is.infinite) 

lang.corr = cor(d.lang, use = "pairwise.complete")
abs(lang.corr) > CORR_THRESHOLD

# include all language variables
```

PCA
```{r PCA with languages variables}
# do pca
pca.lang = prcomp(na.omit(d.lang), scale = TRUE)
barplot(pca.lang$sdev/pca.lang$sdev[1])

# look at variance explained
fviz_eig(pca.lang, addlabels = TRUE, 
               linecolor ="red", geom = "line") +
  theme_bw()
pca.lang = prcomp(na.omit(d.lang), scale = TRUE, tol = .7)

# plot loadings
pcs.lang <- cbind(names(d.lang),
                    gather(as.data.frame(pca.lang$rotation), 
                           pc, value))
names(pcs.lang)[1] = "variable"
 
#pcs.lang$variable =  factor(pcs.lang$variable , levels = pcs.lang$variable[1:8]) # order variables
pcs.lang$magnitude = ifelse(abs(pcs.lang$value)>.2, TRUE, FALSE)
ggplot(pcs.lang) +
  geom_bar(aes(x = variable, y = value,
               fill = magnitude), stat = "identity") +
  facet_wrap(~pc) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  + 
       scale_fill_manual(name = "Magnitude", 
                       values = c("grey","red1")) +
  ggtitle("PCA loadings")

# merge PCAs with full dataset for model fits
d.lang[!is.na(d.lang)] = 0
d.lang[is.na(d.lang)] = 1
good_is = which(rowSums(d.lang) == 0)
d.pca.lang = d.clean[good_is,] %>% cbind(pca.lang$x)

names(d.pca.lang)[names(d.pca.lang) == "PC1"] = "PC1_L"
names(d.pca.lang)[names(d.pca.lang) == "PC2"] = "PC2_L"
```

Model fits using language PCA with all demographics variables
```{r PCA with language variables (lmer), fig.width=12, fig.height = 6}
lang_vars_crit2 = c("PC1_L", "PC2_L")

demo_vars_crit2 = c("n.neighbors_log",
              "mean.temp", "sum.precip", "sd.temp",
              "pop_log", "area_log", "n.neighbors_log",
              "sd.precip_log", "ratio.L2.L1_log")

d.scatter.fam = d.pca.lang %>% 
  select(ratio.L2.L1_log, pop_log,
         n.neighbors_log, area_log, mean.temp, sd.temp, sum.precip, 
         PC1_L, PC2_L, lang.family, native.country) %>%
  gather(lang_measure, lang_value, 
         which(is.element(names(.), lang_vars_crit2))) %>%
  group_by(lang_measure) %>%
  gather(pop_measure, pop_value,
         which(is.element(names(.), demo_vars_crit2)))

is.na(d.scatter.fam) <- sapply(d.scatter.fam, is.infinite) 
d.scatter.fam <- filter(d.scatter.fam, !is.na(pop_value),
                        !is.na(lang_value))

# get model fits
d.model.fits = d.scatter.fam %>%
  group_by(lang_measure, pop_measure) %>%
  do(tidy(lmer(lang_value ~ pop_value +
                 (1|native.country), data=.))) 

line.fits = d.model.fits %>%
          filter(term == "pop_value" | term == "(Intercept)") %>%
  select(-std.error, -statistic, -group) %>%
  spread(term, estimate) %>%
  rename(s = pop_value, 
               i = `(Intercept)`)

sigs = d.model.fits %>%
  filter(term == "pop_value") %>%
  mutate(sig.col = ifelse(statistic > 1.96, "pos",
                          ifelse(statistic < -1.96, "neg",
                                 "none"))) %>%
  mutate(pop_value = .1, lang_value = .1) %>% # this is a hack
  ungroup

ggplot(d.scatter.fam, aes(x = pop_value, y = lang_value)) +
  geom_rect(data = sigs, aes(fill = sig.col), 
          xmin = -Inf, xmax = Inf,
          ymin = -Inf, ymax = Inf, alpha = 0.2) +
  geom_point(size = .3) +
  geom_abline(line.fits, mapping = aes(slope = s, intercept = i), 
              size = 1, color = "green") +
  facet_grid(lang_measure~pop_measure, scales = "free") +
  scale_fill_manual(values = c( "mediumblue", "grey99","red1")) +
  theme_bw() +
  xlab("Demographic variables") +
  ylab("Language variables") +
  theme(legend.position = "none")
```

### All PCAs
Model fits
```{r, fig.width=10, fig.height = 6}
all.pca = left_join(d.pca.demo,
                    d.pca.lang[,c("PC1_L", "PC2_L", "ISO")],
                    by = "ISO")

demo_vars_crit2 = c("hot", "small")
lang_vars_crit2 = c("PC1_L", "PC2_L")

d.scatter.fam = all.pca %>% 
  select(PC1_L, PC2_L, hot, small, 
         lang.family, native.country, ISO, language) %>%
  gather(lang_measure, lang_value, 
         which(is.element(names(.), lang_vars_crit2))) %>%
  group_by(lang_measure) %>%
  gather(pop_measure, pop_value,
         which(is.element(names(.), demo_vars_crit2)))

is.na(d.scatter.fam) <- sapply(d.scatter.fam, is.infinite) 
d.scatter.fam <- filter(d.scatter.fam, !is.na(pop_value),
                        !is.na(lang_value))

# get model fits
d.model.fits = d.scatter.fam %>%
  group_by(lang_measure, pop_measure) %>%
  do(tidy(lmer(lang_value ~ pop_value + (pop_value|lang.family) +
                 (1|native.country), data=.))) 

line.fits = d.model.fits %>%
          filter(term == "pop_value" | term == "(Intercept)") %>%
  select(-std.error, -statistic, -group) %>%
  spread(term, estimate) %>%
  rename(s = pop_value, 
         i = `(Intercept)`)

sigs = d.model.fits %>%
  filter(term == "pop_value") %>%
  mutate(sig.col = ifelse(statistic > 1.96, "pos",
                          ifelse(statistic < -1.96, "neg",
                                 "none"))) %>%
  mutate(pop_value = .1, lang_value = .1) %>% # this is a hack
  ungroup

ggplot(d.scatter.fam, aes(x = pop_value, y = lang_value)) +
  geom_rect(data = sigs, aes(fill = sig.col),
          xmin = -Inf, xmax = Inf,
          ymin = -Inf, ymax = Inf, alpha = 0.2) +
  geom_point(size = .3) +
  geom_abline(line.fits, mapping = aes(slope = s, intercept = i), size = 1, color = "green") +
  facet_grid(lang_measure~pop_measure, scales = "free") +
  scale_fill_manual(values = c( "mediumblue", "grey99","red1")) +
  theme_bw() +
  xlab("Demographic variables") +
  ylab("Language variables") +
  theme(legend.position = "none")
```

Paper plot #2
```{r, eval = F}
dcopy = d.scatter.fam
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)

#Extract Legend 
g_legend<-function(a.gplot){ 
  tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
  legend <- tmp$grobs[[leg]] 
  return(legend)} 

facet_env_names <- c(PC1="PC1: Hot and rainy", PC2="PC2: Small")
facet_lang_names <- c(PC1="PC1: Complex", PC2="PC2: Short words")

pc.demo = ggplot(pcs.demo) +
  geom_bar(aes(x = variable, y = value, fill = variable), stat="identity") +
  facet_wrap(~pc, labeller = labeller(pc = facet_env_names)) +
  ylab("Component loading")+
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        strip.text= element_text(size = 21), 
        axis.title = element_text(size = 21),
        strip.background = element_rect(fill = "grey", colour = "white")) +
  ylim(-.6, .7) +
  scale_x_discrete(breaks=NULL) +
  scale_fill_brewer(palette="YlOrRd")  +
  theme(legend.position = "none")

# reorder factor levels for so organize neatly for PCAS
pcs.lang$variable<- factor(pcs.lang$variable,
                                    levels = rev(c("morphological.complexity",
                                             "scaled.LDT.H", "mean.length",
                                            "n.consonants_log","n.vowels_log")))


pc.lang = ggplot(pcs.lang) +
  geom_bar(aes(x = variable, y = value,
               fill = variable), stat="identity") +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        strip.text= element_text(size = 21), 
        axis.title = element_text(size = 21), 
        strip.background = element_rect(fill = "grey", colour = "white")) +
  coord_flip() +
  facet_grid(pc ~ ., scales="free_x", 
             labeller = labeller(pc = facet_lang_names)) +
  ylim(-.6, .7) +
  scale_x_discrete(breaks=NULL) +
  ylab("Component loading") +
  theme(legend.position = "none")  +
  scale_fill_brewer(palette="BuGn") 

## add missing names:
extra_names = data.frame(ISO = c("tgl", "tam", "sna", "wol"), 
                         language = c("tagalog", "tamil", "shona", "wolof"))

d.scatter.fam = left_join(d.scatter.fam, extra_names, by="ISO") %>%
                  mutate(language = as.factor(ifelse(is.na(language.x),
                                           as.character(language.y),
                                           as.character(language.x))))

p1 <- ggplot(d.scatter.fam, aes(x = pop_value, y = lang_value)) +
  geom_rect(data = sigs, aes(fill = sig.col),
            xmin = -Inf, xmax = Inf,
            ymin = -Inf, ymax = Inf, alpha = 0.2) +

  geom_abline(line.fits, mapping = aes(slope = s, intercept = i), 
              size = 3, color = "grey48" ) +
  facet_grid(lang_measure~pop_measure, scales = "free") +
  geom_text_repel(aes(label = language), size = 5.5) +
  geom_point() +
  #geom_text(aes(label = language), size = 3) +
  scale_fill_manual(values = c( "mediumblue", "grey99","red1")) +
  theme_bw() + 
  theme(strip.background = element_blank(),
        strip.text = element_blank(), 
        axis.title = element_text(size = 24)) +
  ylab("Rotated language values")+
  xlab("Rotated environmental values")+
  theme(legend.position = "none") 
p1

SIZE = 1.6
FONT_SIZE = 20
pc.demo.legend = ggplot(pcs.demo) +
  geom_bar(aes(x = variable, y = value, fill = variable), stat = "identity") +
  scale_fill_brewer(guide = guide_legend(title = NULL, reverse = T, 
                                         keywidth = SIZE, keyheight = SIZE),
                                         palette="YlOrRd") +
  theme(legend.text=element_text(size=FONT_SIZE))
legend.demo <- g_legend(pc.demo.legend) 

pc.lang.legend = ggplot(pcs.lang) +
  geom_bar(aes(x = variable, y = value, fill = variable), stat = "identity") +
  #theme(legend.title=element_blank()) +
  scale_fill_brewer(guide = guide_legend(title = NULL, reverse = T, 
                                         keywidth = SIZE, keyheight = SIZE),
                      palette="BuGn")  +
  theme(legend.text=element_text(size=FONT_SIZE))
legend.lang <- g_legend(pc.lang.legend) 


pdf("../papers/cogsci2016/figs/plot2.pdf", width = 18, height = 12)
#quartz(width = 18, height = 12)
legend.demo$vp$x <- unit(.75, 'npc')
legend.demo$vp$y <- unit(.84, 'npc')
legend.lang$vp$x <- unit(.9, 'npc')
legend.lang$vp$y <- unit(.8, 'npc')
grid.newpage()
pushViewport(viewport(layout = grid.layout(6, 6))) 
print(pc.demo, vp=vplayout(1:2.3,1:4)) 
print(pc.lang, vp=vplayout(3:6,5:6)) 
print(p1, vp=vplayout(3:6,1:4)) 
#grid.draw(legend.demo) 
#grid.draw(legend.lang) 
dev.off()
```


## AOA
### AOA and wordbank data

Grand means across languages
```{r}
# read in word bank data
aoa_data.wb = read.csv("../data/frank_underreview.csv") %>%
  filter(aoa > 5 & aoa < 31)
 
# get common words across languages in wb
all_words = levels(aoa_data.wb$uni_lemma)
for (i in 1:length(levels(aoa_data.wb$language))){
  this_language = aoa_data.wb[aoa_data.wb$language == levels(aoa_data.wb$language)[i],]
  all_words = intersect(all_words, unique(this_language$uni_lemma))
}

aoa.means.all = aoa_data.wb %>% 
  mutate(language = tolower(language)) %>%
  filter(is.element(uni_lemma, all_words)) %>%
  group_by(language) %>%
  summarise(mean.aoa.wb = mean(aoa)) %>%
  left_join(select(d.clean, mean.aoa, language)) %>%
  rename(mean.aoa.lu = mean.aoa) 

ggplot(aoa.means.all, aes(x = mean.aoa.lu, y = mean.aoa.wb)) +
  geom_text(aes(label = language)) +
  geom_smooth(method = "lm") +
  ggtitle("Grand means across langauges (wb common words only; n = 117)") +
  theme_bw()

cor.test(aoa.means.all$mean.aoa.wb, 
         aoa.means.all$mean.aoa.lu, 
         use = "pairwise.complete")
```

Now lets look at the same words for both data sets [n = 56]
```{r}
all.wb.words = unlist(lapply(strsplit(
    as.character(all_words)," "), function(x) x[1]))

aoa.raw$word  = unlist(lapply(strsplit(
    as.character(aoa.raw$word)," "), function(x) x[1]))

aoa.lu.common.words.only = aoa.raw %>%
  #filter lu words based on the ones that are common in wb
  filter(is.element(word,all.wb.words)) %>% 
  gather(language_aoa, aoa, grep("aoa", 
                                 names(aoa.raw))) %>%
  mutate(language_aoa = tolower(unlist(lapply(strsplit(
    as.character(language_aoa),"_"), function(x) x[1])))) %>%
  select(-3:-27) %>%
  rename(language = language_aoa) %>%
  group_by(language) %>%
  summarize(mean.aoa = mean(aoa, na.rm = T)) 

aoa.means.all.common = aoa_data.wb %>% 
  mutate(language = tolower(language)) %>%
  filter(is.element(uni_lemma, all_words)) %>%
  group_by(language) %>%
  summarise(mean.aoa.wb = mean(aoa)) %>%
  left_join(select(aoa.lu.common.words.only, mean.aoa, language)) %>%
  rename(mean.aoa.lu = mean.aoa) 

ggplot(aoa.means.all.common, aes(x = mean.aoa.lu, y = mean.aoa.wb)) +
  geom_text(aes(label = language)) +
  geom_smooth(method = "lm") +
  ggtitle("Grand means across langauges (common words only in both datasets; n = 56)") +
  theme_bw()

cor.test(aoa.means.all.common$mean.aoa.wb, 
         aoa.means.all.common$mean.aoa.lu, 
         use = "pairwise.complete")
```

At word level, by language
```{r,  fig.width=10, fig.height = 3}
aoa.lu.common.words.only = aoa.raw %>%
  filter(is.element(word,all.wb.words))  %>% #filter lu words based on the ones that are common in wb
  gather(language_aoa, aoa, grep("aoa", 
                                 names(aoa.raw))) %>%
  mutate(language_aoa = tolower(unlist(lapply(strsplit(
    as.character(language_aoa),"_"), function(x) x[1])))) %>%
  select(-3:-27) %>%
  rename(language = language_aoa) %>%
  select(word, language, aoa) %>%
  group_by(word, language) %>%
  summarise(aoa = mean(aoa, na.rm = T)) %>%
  mutate(dataset = "lu")

aoa_data.wb$uni_lemma = unlist(lapply(strsplit(
    as.character(aoa_data.wb$uni_lemma)," "), function(x) x[1]))

common_words_across_all_langs_and_datasets = unique(aoa.lu.common.words.only$word)

aoa.wb.common.words.only = aoa_data.wb %>% 
  mutate(language = tolower(language)) %>%
  group_by(language, uni_lemma) %>%
  summarise(aoa = mean(aoa, na.rm = T)) %>%
  filter(is.element(uni_lemma, common_words_across_all_langs_and_datasets)) %>%
  rename(word = uni_lemma) %>%
  select(language, word, aoa) %>%
  mutate(dataset = "wb") 

common_languages = intersect(tolower(as.character(unique(aoa_data.wb$language))), #wb
                             unique(aoa.lu.common.words.only$language)) #lu

all.aoa.by.word = rbind(aoa.wb.common.words.only, 
                        aoa.lu.common.words.only) %>%
  filter(is.element(language, common_languages))  %>%
  group_by(language, dataset) %>%
  spread(dataset, aoa)

ggplot(all.aoa.by.word, aes(x = lu, y = wb, 
                            group = language, color = language)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(.~language) +
  ggtitle("By word means for each language (common words only (n = 56))") +
  theme_bw() +
  theme(legend.position="none") 

summary(lm(wb ~ lu + language, data = all.aoa.by.word))
summary(lmer(wb ~ lu + (lu|language), all.aoa.by.word))
```

By common words only
```{r}
all.means.by.word = all.aoa.by.word %>%
  group_by(word) %>% 
  summarise(mean.lu = mean(lu, na.rm = T),
             mean.wb = mean(wb, na.rm  = T))

ggplot(all.means.by.word, 
       aes(x = mean.lu, y = mean.wb)) +
  geom_text(aes(label = word)) +
  geom_smooth(method = "lm") +
  ggtitle("By word means (common words only (n = 56)") +
  theme_bw()

cor.test(all.means.by.word$mean.lu, all.means.by.word$mean.wb)
```

For the 7 common languages, WB and Lu are relatively correlated (r = .36).

### AOA and environmental variables 
```{r fig.width=10, fig.height = 3}
d.scatter.fam = d.clean %>% 
  select(pop_log, ratio.L2.L1_log,
         n.neighbors_log, area_log, mean.temp, sd.temp, sum.precip,
         sd.precip_log, language, mean.aoa, lang.family, native.country) %>%
  gather(pop_measure, pop_value,
         which(is.element(names(.), demo_vars_crit)))

d.model.fits = d.scatter.fam %>%
  group_by(pop_measure) %>%
  do(tidy(lmer(mean.aoa ~ pop_value + (pop_value|lang.family) +
                 (1|native.country), data=.))) 

line.fits = d.model.fits %>%
        filter(term == "pop_value" | term == "(Intercept)") %>%
        select(-std.error, -statistic, -group) %>%
        spread(term, estimate) %>%
        rename(s = pop_value, 
               i = `(Intercept)`)

sigs = d.model.fits %>%
  filter(term == "pop_value") %>%
  mutate(sig.col = ifelse(statistic > 1.96, "pos",
                          ifelse(statistic < -1.96, "neg",
                                 "none"))) %>%
  mutate(pop_value = .1, mean.aoa = .1) %>% # this is a hack
  ungroup

ggplot(d.scatter.fam, aes(x = pop_value, y = mean.aoa)) +
  geom_rect(data = sigs, aes(fill = sig.col),
            xmin = -Inf, xmax = Inf,
            ymin = -Inf, ymax = Inf, alpha = 0.2) +
  geom_point() +
  geom_abline(line.fits, mapping = aes(slope = s, intercept = i), 
              size = 2) +
  facet_grid(.~pop_measure, scales = "free",
                          labeller = labeller(pop_measure = facet_pop_names)) +
  geom_text_repel(aes(label = language), size = 3) +
  scale_fill_manual(values = c( "mediumblue", "grey99","red1")) +
  theme_bw() + 
  theme(legend.position = "none") 
```

Temperature is positively correlated with AOA, and precipitation is negatively correlated. Children learn words later in hotter places with little variability in preicipitation.

### AOA and language variables 
```{r AOA and language variables, fig.width=10, fig.height = 3}
lang_vars_crit = c("p.complexity.bias","TTR.LDT",
                   "mean.dependency.length", "mean.length",
                   "n.consonants_log", "n.vowels_log",
                   "morphological.complexity")

d.scatter.fam = d.clean %>% 
  filter(language != "xhosa") %>% # an outlier
  select(p.complexity.bias,TTR.LDT,
                   mean.dependency.length, mean.length,
                   n.consonants_log, n.vowels_log,
                   morphological.complexity, mean.aoa, 
                   lang.family, native.country, language) %>%
  gather(lang_measure, lang_value,
         which(is.element(names(.), lang_vars_crit)))

# get model fits
d.model.fits = d.scatter.fam %>%
  group_by(lang_measure) %>%
  do(tidy(lmer(mean.aoa ~ lang_value + (lang_value|lang.family) +
                 (1|native.country), data=.))) 

line.fits = d.model.fits %>%
        filter(term == "lang_value" | term == "(Intercept)") %>%
        select(-std.error, -statistic, -group) %>%
        spread(term, estimate) %>%
        rename(s = lang_value, 
               i = `(Intercept)`)

sigs = d.model.fits %>%
  filter(term == "lang_value") %>%
  mutate(sig.col = ifelse(statistic > 1.96, "pos",
                          ifelse(statistic < -1.96, "neg",
                                 "none"))) %>%
  mutate(lang_value = .1, mean.aoa = .1) %>% # this is a hack
  ungroup

ggplot(d.scatter.fam, aes(x = lang_value, y = mean.aoa)) +
  geom_rect(data = sigs, aes(fill = sig.col),
           xmin = -Inf, xmax = Inf,
           ymin = -Inf, ymax = Inf, alpha = 0.2) +
  geom_point() +
  geom_abline(line.fits, mapping = aes(slope = s, intercept = i), 
              size = 2) +
  facet_grid(.~lang_measure, scales = "free") +
  geom_text_repel(aes(label = language), size = 3) +
  scale_fill_manual(values = c( "grey99", "red1","mediumblue")) +
  theme_bw() + 
  theme(legend.position = "none") 

## TTR significant in correlation
cor.test(d.clean$TTR.LDT, d.clean$mean.aoa)
```

Predicted relationship with AOA number of consonants (TTR has reliable correlation, but not significant in lmer)

#### AOA and PCA
```{r}

```

### AOA misc. morphosyntactic features

#### AOA and word order
```{r}
d.clean = ld %>%
  select(ISO, DRYSOV) %>%
  left_join(d.clean, by="ISO")

old = 1:7
new = new = c("SOV", "SVO", "VSO", "VOS", "OVS", "OSV", "none")
d.clean$DRYSOV = mapvalues(d.clean$DRYSOV, 
                         from = as.character(old),
                         to = as.character(new))

d.clean %>%
  distinct(ISO) %>%
  filter(!is.na(mean.aoa))%>%
  group_by(DRYSOV) %>%
  multi_boot_standard("mean.aoa", na.rm = T) %>%
  ggplot(aes(x = DRYSOV, y = mean, fill = DRYSOV)) +
    geom_bar(stat = "identity") +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
                  size=.3,
                  width=.2,
                  position=position_dodge(.9)) +
  theme_bw()
```

Too small of a sample to see differences across categories.

#### AOA and isolating vs. concatenating
```{r}
d.clean = ld %>%
  select(ISO, BICFUS) %>%
  rename(affix.type = BICFUS) %>%
  left_join(d.clean, by="ISO") %>%
  mutate(affix.type = as.factor(affix.type))
summary(d.clean$affix.type)
old = 1:7
new = new = c("concatenative", "isolating", "tonal",
              "isolating", "concatenative", "concatenative", "both")

d.clean$affix.type = mapvalues(d.clean$affix.type, 
                         from = as.character(old),
                         to = as.character(new))

d.clean %>%
 filter(!is.na(mean.aoa), !is.na(affix.type)) %>%
  group_by(ISO) %>%
  summarise(n = n(),
            affix.type = affix.type[1]) 
```
All are concatenative that we have data for (n = 9 for AOA ^ WALS). Can't do this analysis.

#### AOA and prefixing vs. suffixing
```{r}
d.clean = ld %>%
  select(ISO, DRYPRE) %>%
  rename(preVsuf = DRYPRE) %>%
  left_join(d.clean, by="ISO") %>%
  mutate(preVsuf = as.factor(preVsuf))
summary(d.clean$preVsuf)
old = 1:6
new = c("none", "suffix", "suffix",
              "prefix", "prefix", "prefix")

d.clean$preVsuf = mapvalues(d.clean$preVsuf, 
                         from = as.character(old),
                         to = as.character(new))

d.clean %>%
 filter(!is.na(mean.aoa), !is.na(preVsuf)) %>%
  group_by(ISO) %>%
  summarise(n = n(),
            affix.type = preVsuf[1]) %>%
  select(-n) %>%
  summary()
```

All but two are suffixing. Can't do this analysis either.

